---
title: "Reproducible Research Assess 1"
author: "Clare Wilkinson"
date: "Wednesday, March 11, 2015"
output: pdf_document
---

Reproducible Research Assessment 1
======================================

Set the working directory (note this will be different for everyone)
```{r, echo=TRUE}
setwd("~/PhD/Statistics/Cousera courses/Reproducible research/Assessment 1/repdata_data_activity")
```

Load the data
```{r, echo=TRUE}
activity_raw <- read.csv('activity.csv', , stringsAsFactors=FALSE)
```

Process the data for your analysis
```{r, echo=TRUE}
# Transform the date variable to an actual date format
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")

# Compute the weekdays from the date variable
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)

# Compute the type of day (weekend or weekday)
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                     activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

# Create the final data.frame
activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

# Remove the raw data set and keep the processed data
rm(activity_raw)

#Display top of table 
head(activity)
```

What is mean total number of steps taken per day? (Ignore the missing values)
--------------------------------------------------

1. The number of steps taken per day
```{r, echo=TRUE}
# the total number of steps each day (NA values removed)
sumdata <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)

# Rename variable
names(sumdata) <- c("date", "totalsteps")
```

2. a histogram of the total number of steps taken each day
```{r, echo=TRUE}
hist(sumdata$totalsteps,  breaks=seq(from=0, to=25000, by=2000), ylim=c(0,20), col="red", xlab="Total number of steps",  main="Histogram   of the total number of steps taken each day")
```

3. The mean and median of the total number of steps taken per day
```{r, echo=TRUE}
mean(sumdata$totalsteps)
median(sumdata$totalsteps)
```


What is the average daily activity pattern?
-------------------------------------------
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE}
#Create the new dataset of mean steps for each interval
meandata <- aggregate(activity$steps, by=list(activity$interval),  FUN=mean, na.rm=TRUE)

# Rename the variables
names(meandata) <- c("interval", "mean")

# Plot the time series
plot(meandata$interval, meandata$mean, type="l", col="red", xlab="Interval (minutes)", ylab="Average number of steps", 
      main="Time-series of the average number of steps per intervals")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
# We find the position of the maximum mean
maxposition <- which(meandata$mean == max(meandata$mean))

# We lookup the value of interval at this position
maxinterval <- meandata[maxposition, 1]
```


Imputing missing values
-------------------------

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo=TRUE}
NAcount <- sum(is.na(activity$steps))
```

2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be 
sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r, echo=TRUE}
# Find the NA positions
napositions <- which(is.na(activity$steps))

#Use the mean number of steps for each interval for filling in the missing values
# Create a vector of means, that is the number of nas in length
meanvectorforNA <- rep(mean(activity$steps, na.rm=TRUE), times=length(napositions))
```

3.Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo=TRUE}
# Replace the NAs by the means
activity[napositions, "steps"] <- meanvectorforNA
```

4.Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

The histogram:
```{r, echo=TRUE}
# Compute the total number of steps each day (NA values removed)
sumdata2 <- aggregate(activity$steps, by=list(activity$date), FUN=sum)

# Rename variables
names(sumdata2) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sumdata2$total, breaks=seq(from=0, to=25000, by=2500),col="red", xlab="Total number of steps", ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")
```
The mean and median: 
```{r, echo=TRUE}
mean(sumdata2$total)
median(sumdata2$total)
```

These values are much larger than the previous estimates, as putting in the missing values, provides more data, so the numbers were expected to increase.


Are there differences in activity patterns between weekdays and weekends?
---------------------------------------------------------------------------

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r, echo=TRUE}
# The factor variable "daytype" is already in the activity data
head(activity)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r, echo=TRUE}
# Load the lattice package
library(lattice)

# calculate average number of steps taken,across all daytype variable
meandata2 <- aggregate(activity$steps, by=list(activity$daytype, activity$weekday, activity$interval), mean)

# Rename variables
names(meandata2) <- c("daytype", "weekday", "interval", "mean")

#The time series plot
xyplot(mean ~ interval | daytype, meandata2, type="l",  lwd=1, xlab="Interval", ylab="Number of steps", layout=c(1,2))
```

